---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# ukbschema

<!-- badges: start -->
<!-- badges: end -->

This R package can be used to create and/or load the [UK Biobank Data Showcase schemas](http://biobank.ctsu.ox.ac.uk/crystal/schema.cgi) for further use.

## Installation

You can install the current version of ukbschema from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("bjcairns/ukbschema")
```
## Examples

```{r prelim, echo=FALSE, include=FALSE}
try(rm(list=c("db","sch")))
```

The package exports two functions. The first is `create_schema_db()`, which downloads the schema tables and saves them to an SQLite database:

```{r create}
library(ukbschema)

db <- create_schema_db(path = tempdir())
```

```{r info, echo=FALSE, include=FALSE}
file <- paste0(tempdir(), "\\ukb-schema-", Sys.Date(), ".sqlite")
fsize <- round(file.info(file)$size/1e6, 1)
```

By default, the database is named `ukb-schema-YYYY-MM-DD.sqlite` and placed in the current working directory (`path = tempdir()` in the above example puts it in the current temporary directory instead). At time of compilation of this README (`r format(Sys.Date(), "%d %b %Y")`), the size of the .sqlite database file produced by `create_schema_db()` is approximately `r fsize`MB.

The second function, `load_schemas()`, loads the tables from the database and stores them as tibbles in a named list:

```{r load}
sch <- load_schemas(db = db)
names(sch)
```

Note that without further arguments, `create_schema_db()` tidies up the database to give it a more consistent relational structure (the changes are summarised in the output of the first example, above). Alternatively the raw data can be loaded with the `as_is` argument:

```{r create_as_is}
db <- create_schema_db(path = tempdir(), overwrite = TRUE, as_is = TRUE)
```

## Why R?

This package was originally written in bash (a Unix shell scripting language). However, R is more accessible and all dependencies are loaded when you install the package; there is no need to install any secondary software (not even SQLite).

## Notes

#### Design notes

* All the encoding value tables (`esimpint`, `esimpstring`, `esimpreal`, `esimpdate`, `ehierint`, `ehierstring`) have been harmonised and combined into a single table `encvalues`. The `value` column in `encvalues` has type `TEXT`, but a `type` column has been added in case the value is not clear from context. The original type-specific tables have been deleted.
* To avoid redunancy, category parent-child relationships have been moved to table `categories`, as column `parent_id`, from table `catbrowse` (which has been deleted).
* The value types described [on the UKB Showcase](http://biobank.ctsu.ox.ac.uk/crystal/help.cgi?cd=value_type) have been added manually to a table `valuetypes` and appropriate ID references have been renamed to `value_type_id` in tables `fields` and `encodings`.

#### Known code issues

* The UK Biobank database schema are regularly updated as new data are added to the system. ukbschema does not currently include a facility for updating the database; it is necessary at present to create a new database.
* Any [other issues](https://github.com/bjcairns/ukbschema/issues).

#### Known data issues

* Category 119 does not exist in the data file for categories, but is referenced in table `catbrowse`. As a work-around it has been manually added to the database (if `as_is = FALSE` in `create_schema_db()`) by inferring information about it from related category [100032](https://biobank.ctsu.ox.ac.uk/crystal/label.cgi?id=100032).
* Value types as described in http://biobank.ctsu.ox.ac.uk/crystal/help.cgi?cd=value_type are not included in the UKB schema. See [Design notes](https://github.com/bjcairns/ukbschema#design-notes), above.

