---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# ukbschemas

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Build Status](https://travis-ci.com/bjcairns/ukbschemas.svg?token=tA2cYTLpigx5VuTgcHFd&branch=master)](https://travis-ci.com/bjcairns/ukbschemas)
<!-- badges: end -->

This R package can be used to create and/or load the [UK Biobank Data Showcase schemas](http://biobank.ctsu.ox.ac.uk/crystal/schema.cgi) for further use.

## Installation

You can install the current version of ukbschemas from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("bjcairns/ukbschemas")
```
## Examples

```{r prelim, echo=FALSE, include=FALSE}
try(rm(list=c("db","sch")))
```

The package supports two workflows. The recommended approach is to use `ukbschemas_db()` to download the schema tables and save them to an SQLite database:

```{r create}
library(ukbschemas)

db <- ukbschemas_db(path = tempdir())
```

```{r info, echo=FALSE, include=FALSE}
file <- paste0(tempdir(), "\\ukb-schemas-", Sys.Date(), ".sqlite")
fsize <- round(file.info(file)$size/1e6, 1)
```

By default, the database is named `ukb-schemas-YYYY-MM-DD.sqlite` (where `YYYY-MM-DD` is the current date) and placed in the current working directory. (`path = tempdir()` in the above example puts it in the current temporary directory instead.) At the most recent compilation of this README (`r format(Sys.Date(), "%d %b %Y")`), the size of the .sqlite database file produced by `ukbschemas_db()` was approximately `r fsize`MB.

To retrieve the schemas from the database, use `load_db()`, which loads the tables from the database and stores them as tibbles in a named list:

```{r load}
sch <- load_db(db = db)
names(sch)
```

Note that without further arguments, `ukbschemas_db()` tidies up the database to give it a more consistent relational structure (the changes are summarised in the output of the first example, above). Alternatively the raw data can be loaded with the `as_is` argument:

```{r create_as_is}
db <- ukbschemas_db(path = tempdir(), overwrite = TRUE, as_is = TRUE)
```

The second approach is to download the schemas and store them in memory in a list, and save them to a database only as requried. This is not recommended, because it is better (for everyone) not to download the schema files every time they are needed, and because the the database assumes a certain structure that should be guaranteed when the database is saved. If you still want to take this approach, use:

```{r inmemory}
sch <- ukbschemas()
names(sch)

db <- save_db(sch, path = tempdir(), overwrite = TRUE)
```

## Why R?

This package was originally written in bash (a Unix shell scripting language). However, R is more accessible and all dependencies are loaded when you install the package; there is no need to install any secondary software (not even SQLite).

## Notes

#### Design notes

* All the encoding value tables (`esimpint`, `esimpstring`, `esimpreal`, `esimpdate`, `ehierint`, `ehierstring`) have been harmonised and combined into a single table `encvalues`. The `value` column in `encvalues` has type `TEXT`, but a `type` column has been added in case the value is not clear from context. The original type-specific tables have been deleted.
* To avoid redunancy, category parent-child relationships have been moved to table `categories`, as column `parent_id`, from table `catbrowse` (which has been deleted).
* Reference to the category to which a field belongs is in the `main_category` column in the `fields` schema, but has been renamed to `category_id` for consistency with the `categories` schema.
* The value types described [on the UKB Showcase](http://biobank.ctsu.ox.ac.uk/crystal/help.cgi?cd=value_type) have been added manually to a table `valuetypes` and appropriate ID references have been renamed to `value_type_id` in tables `fields` and `encodings`.

#### Known code issues

* The UK Biobank data schemas are regularly updated as new data are added to the system. ukbschemas does not currently include a facility for updating the database; it is necessary to create a new database.
* Any [other issues](https://github.com/bjcairns/ukbschemas/issues).

